evidence of adaptation-independent time-warp spike timing noise correlations in cortical networks

### 2020-08-12 manuscript.tex
End of day!

The neural coding problem addresses the question of how information is encoded by patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}. Whilst precise temporal spike patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained. Such variability leads to the common assumption that the brain uses a population code based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}. Previous studies have shown that spike count variability is shared across the population in early sensory cortices \cite{scholvinck2015cortical, lin2015nature}. Single cortical neurons can generate highly reliable spike trains to direct stimulation in vitro \cite{mainen1995reliability}. In vivo shared variability is therefore likely to arise from variability in synaptic inputs, reflecting cortical network dynamics \cite{carandini2004amplification}.

In early visual cortex, trial-to-trial spike count variability is correlated across the population and is modulated by two global factors, which fluctuate the level of excitability from trial-to-trial \cite{lin2015nature}. The first of these is an additive factor, which is explained by the interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}. The second is a multiplicative factor, which also modulates the trial-to-trial fluctuation in excitability across the population, and has been proposed to be caused by top-down feedback from higher cortices \cite{lin2015nature}. Fluctuations in spontaneous activity have also been shown to gate sensory-evoked responses \cite{luczak2013gating}.

Fluctuations in trial-to-trial excitability are referred to herein as trial-to-trial fluctuations in continuous cortical state \cite{scholvinck2015cortical}, which may be multi-dimensional and can be separated into short and long-timescale changes. In this paper, long-timescale fluctuations refer to large changes taking place over multiple trials caused by processes such as neural adaptation in response to repetitive stimulation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, or changes in the level of anaesthesia. Short-timescale fluctuations refer to random trial-to-trial fluctuations in excitability around the current general level of excitability.

@@

This study aims to characterise the non-trivial nature of trial-to-trial spike time (rather than spike count) noise correlations, as a component step towards investigating the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state. 
% which would introduce the possibility of state-dependent encoding and decoding (see below). 
The characterisation controls for longer timescale changes in excitability to estimate how stimulus representations are modulated by trial-to-trial fluctuations in state at a given point in time. 
% This offers a novel prediction of the encoding capacity under a state-dependent spike time encoding regime. 
Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.

%  low 2-dimensional continuous latent state .

% It is possible that spike time variability is caused and modulated across a population by such a low-dimensional continuous latent state.

@@

Responses following sensory stimulation can be analysed to explore such principles of neural coding. 
% Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
A multi-neuron response can be expressed as a point in space, where each dimension represents a spike train metric. Responses to multiple stimulus presentations form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig. 1a).
% Information theoretic approaches aim to understand 
If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a).

@@

@@

Neural coding classically considers such response-distributions with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}. `Noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimuli, whilst `signal correlations' concern tuning similarity between neurons. With respect to fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.

@@

It is important to consider that if downstream neurons are also modulated by the same population-wide short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
When the fluctuating state has a value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig. 1c). State-dependent decoding would therefore offer an improved encoding capacity, compared to the classical fixed-state decoder interpretation.

@@

Characterising noise correlations is complicated by the heterogeneity of sample response distribution structure and long-timescale changes 
% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
% Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also 
which, could cause correlative structure in the stimulus-response distribution (Fig. X), and must be controlled for when characterising `adaptation-independent' noise correlations.

@@

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of short-timescale spike time noise correlations for first-spike patterns, by controlling for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
% towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state. The classical `stimulus-dependency' of noise correlations is left to future studies. 
A previous study analysing the same data set found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}. Another study analysing the data set could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}. 
% The spike time noise correlations uncovered in this study 
% offer a new opportunity to study population-wide modulation by low-dimensional latent states and 
% controls for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
% and was confirmed to occur for 7 out of 8 neuron groups in this study.

@@

For repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked. For example, Fig. X shows first spike pairs for a pair of units (L5AI and L4I) in the principle column in response to a 4Hz stimulation of the corresponding principle whisker. First spike pairs appear to come from one of two clusters (blue and green dots) for presentations of the same stimulus. Moreover, the green cluster appears correlated with non-unitary gradient (different from $45^{\circ}$).

Correlated clusters were observed in a large number of first spike pair distributions (Supplementary Vid. 1). A novel algorithm extracted isolated clusters from single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}. It was also shown, however, that first spike pairs within these isolated clusters were positively correlated above chance-levels. These correlated clusters represent reliable regions of response distribution space that allow analysis of short- and long-timescale spike timing noise correlations. A large proportion of these noise correlations would have been missed without clustering.

% The paper aims to test that the occurrence of correlated clusters is above chance and to test the degree to which such correlations are caused by adaptative and non-adaptative processes.

% (Fig X; see plots in the top left of Supplementary Vid. 1).

Interestingly, many of the correlated clusters had non-unitary correlation angles. 
Importantly, many clusters remained correlated with non-unitary gradient after controlling for longer-timescale neural adaptation,
% using time series analysis, 
indicative of short-timescale fluctuating state, which varies the latency of different neurons at a different rate.
% giving evidence of precise non-adaptive noise correlations
% Such non-unitary spike time correlation angles cannot be 
This cannot explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle).

This novel finding of `adaptation-independent clustered time-warp spike-timing noise correlations' shows that the distribution of decodable first spike time relative differences for pairs of neurons at a point in time are `warped' from trial-to-trial by some short-timescale fluctuating continuous cortical state.
% This phenomenon is referred to as `adaptation-independent clustered time-warped spike-timing noise correlations'. 
This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not simultaneously consider stimulus encoding, clustering, adaptation and different warping functions for different neurons \cite{williams2020discovering}.

The characterisation of adaptation-independent clustered spike-timing noise correlations is useful under the fixed-state decoder interpretations. More interestingly, these noise correlations open the possibility of state-dependent spike time coding. Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster, giving a prediction of the capacity of state-dependent encoding, which surpasses previously proposed coding forms.

Adaptation-independent non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers. This offers suggestive evidence of a low-dimensional fluctuating state affecting the spike times of spatially distanced neurons, which in turn could modulate downstream neurons and cause them to respond in a state-dependent manner. 
% The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. 
Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by low trial overlap between clusters. This paper provides
% in extracting and characterising adaptation-independent correlated clusters provides 
a useful step towards investigating this and shows that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.
% confirming or rejecting state-dependent spike-time coding.

@@

% Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, the paper aims to address the following questions:

@@

Analysing the joint responses of two or more units to repeated presentations of a single stimulus offers the opportunity to discern noise correlations caused by adaptative or non-adaptative processes.
A two-stage algorithm was developed to quantify the occurrence and angle of correlated clusters observed in first spike response distributions for pairs of units (Fig. X; top left of Supplementary Vid. 1). Stage 1 of the algorithm extracted well isolated clusters defined by the $4\sigma$ boundary of flat 2D Gaussian ellipses (see Methods \ref{methods:flat_clusters}; e.g. flat solid green ellipse in Fig. 3a).
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0}
The algorithm was verified by confirming that the algorithm extracted correlated clusters at chance levels for multiple control data sets (see Methods \ref{methods:correlation_surrogate_control_datasets}; Supplementary Fig. 3). The correlated clusters can be analysed to test whether the correlations are explained by adaptative or non-adaptative processes.

@@

\subsection*{Physiological Relevance}
% It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. 
Testing the state-dependent decoding performance of the clusters is left for future studies, as it would require explicit knowledge of the underlying shared state or an estimate from the spike times of other neurons (which is difficult for the current extracted cluster counts per stimulus in single experimental sessions).
Neuronal subthreshold membrane potentials would be a candidate for such state, as they are correlated between neurons across areas \cite{poulet2008internal} and affect action potential initiation \cite{petersen2003interaction}. The LFP may offer a useful proxy for the subthreshold membrane potential.

@@

\subsection*{Non-stationarity}

Future studies of non-stationary time-warped continua should carefully consider non-stationarity. Adaptation describes the `continuous recalibration of the sensory system to compensate for the changes in the statistics of the input stream' \cite{kheradpezhouh2017response}, with specific functional roles including gain modulation \cite{ahissar2000transformation, heiss2008shift, khatri2009stimulus, mohar2013opposite}, efficiency \cite{sharpee2006adaptive} and detectability of novel stimuli \cite{dragoi2002dynamics, ulanovsky2003processing}. The rate and degree of adaptation depends on stimulus parameters, neuron location and neuron properties (see \cite{kheradpezhouh2017response} for review). Adaptation modulates correlations and response heterogeneity (see \cite{kheradpezhouh2017response} for review), and depends on brain state and anaesthesia \cite{castro2004absence, katz2012trial}. Adaptation from repetitive stimulation changes excitatory-inhibitory balance \cite{higley2006balanced, heiss2008shift, malina2013imbalance} thought to be caused by short term synaptic depression (due to neurotransmitter depletion) observed in trigeminal complex \cite{mohar2013opposite}, thalamus \cite{deschenes2003relay} and cortex \cite{higley2006balanced, heiss2008shift, malina2013imbalance} (mostly at thalamocortical synapses \cite{chung2002short}).

In mouse barrel cortex, spiking statistics change through development \cite{van2017layer}. Functional imaging in adult mice shows day-to-day maintenance of population activity \cite{margolis2012reorganization, mayrhofer2015sparse} but variability in the day-to-day stability of single unit representations \cite{mayrhofer2015sparse}. Functional imaging shows that in addition to tuning sensory maps, plasticity and sensory experience affects response gain, variability, firing correlations, and top-down modulation by task context (see \cite{lemessurier2018plasticity}).

Mechanisms of attention \cite{cohen2010neuronal} and plasticity \cite{lemessurier2018plasticity}.
% Collectively, these studies characterise the barrel cortex as a highly dynamic system, with representations continuously changing.

>>>>

The neural coding problem addresses the question of how information is encoded by patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}. Whilst precise temporal spike patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained. Such variability leads to the common assumption that the brain uses a population code based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}. Previous studies have shown that spike count variability is shared across the population in early sensory cortices \cite{scholvinck2015cortical, lin2015nature}. As single cortical neurons can generate highly reliable spike trains to direct stimulation in vitro \cite{mainen1995reliability}, shared variability in vivo is likely to arise from variability in synaptic inputs reflecting cortical network dynamics \cite{carandini2004amplification}.

In early visual cortex, trial-to-trial spike count variability is correlated across the population and is modulated by two global factors, which fluctuate the level of excitability from trial-to-trial \cite{lin2015nature}. The first of these factors is explained by an additive interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}. The second is a multiplicative factor proposed to be caused by top-down feedback from higher cortices \cite{lin2015nature}. 
Changes in excitability are herein referred to as changes in a continuous cortical state \cite{scholvinck2015cortical}.

% Fluctuations in spontaneous activity also gate sensory-evoked responses \cite{luczak2013gating}.

% which may be multi-dimensional. 
Changes in such cortical state can happen on long or short-timescales, and can be either specific or general across the population (Fig. 0a). In this paper, long-timescale state changes refer to slow changes in neuronal excitability caused by processes such as changes in the level of anaesthesia or neural adaptation \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}. Such processes are more specific than general across the population. Short-timescale state changes refer to random trial-to-trial fluctuations in excitability around the longer-timescale modulated state.

@@

% Study of the modulation of spike times by short-timescale changes in state are limited. 
% Towards investigating whether spike times are modulated by short-timescale fluctuations of a low-dimensional cortical state, this study aims to characterise the nature of spike time noise correlations, whilst controlling for long-timescale changes in excitability.

The paper takes a step towards uncovering the dimensionality of possible population-wide short-timescale state changes, by better characterising pairwise spike time noise correlations, whilst controlling for long-timescale state changes.
% This study aims to characterise the non-trivial nature of trial-to-trial spike time noise correlations, as a component step towards investigating the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state. 
% which would introduce the possibility of state-dependent encoding and decoding (see below). 
% The characterisation controls for longer timescale changes in excitability to estimate how stimulus representations are modulated by trial-to-trial fluctuations in state at a given point in time. 
This offers a novel prediction of the encoding capacity under a state-dependent coding regime. 
Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition. Responses following sensory stimulation can be analysed to explore such principles of neural coding.

%  low 2-dimensional continuous latent state .  
% It is possible that spike time variability is caused and modulated across a population by such a low-dimensional continuous latent state.

@@

% Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
A multi-neuron stimulus response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric. Responses to multiple stimulus presentations form a sample stimulus-response distribution in this space. Neural coding classically considers such response-distributions with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}. All samples from the sample stimulus-response distribution are therefore used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig. 1a).

% Information theoretic approaches aim to understand 
If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a). `Noise correlations' refer to shared trial-to-trial variability over repeated presentations of single stimuli, whilst `signal correlations' concern tuning similarity between neurons. With respect to fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.

@@

It is important to consider that if downstream neurons are also modulated by the same population-wide short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
When the fluctuating state has a value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig. 1c). State-dependent decoding therefore offers an improved coding capacity.
% , compared to the classical fixed-state decoder interpretation.

Long-timescale changes in excitability make it difficult to accurately characterise the stimulus response distribution from which samples are drawn at a point in time. 
% including noise correlation structure caused by short-timescale fluctuations in excitability. 
Long-timescale changes are also likely to cause correlative structure in the stimulus-response distribution that could be misinterpreted as noise correlations caused by a short-timescale fluctuating state. If the effect of long-timescale changes in excitability can be removed or controlled for, it may be possible to give an estimate of the response distribution on a single trial. This could enable characterisation of noise correlations, which in turn could uncover short-timescale fluctuations in a population-wide state. Removing the effect of long-timescale changes and characterising subsequent adaptation-controlled response distributions, is made difficult by heterogeneity in single unit processes of adaptation and the structure of original sample response distributions.

% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
% Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also

@@

@@

@@

@@

Using a large dataset of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of short-timescale spike time noise correlations for first-spike patterns, by controlling for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
% towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state. The classical `stimulus-dependency' of noise correlations is left to future studies. 
A previous study analysing the same dataset found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}. Another study analysing the dataset could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}. 
% The spike time noise correlations uncovered in this study 
% offer a new opportunity to study population-wide modulation by low-dimensional latent states and 
% controls for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
% and was confirmed to occur for 7 out of 8 neuron groups in this study.

@@

% \section{Results}

For repeated trials of a single stimulus condition, the first spike times of two units were plotted against each other for trials on which they both spiked. For example, Fig. X shows first spike pairs for a pair of units (L5AI and L4I) in the principle column in response to a 4Hz stimulation of the corresponding principle whisker. First spike pairs appear to come from one of two clusters (blue and green dots) for presentations of the same stimulus condition. Moreover, the green cluster appears correlated.

Correlated clusters were observed in a large number of first spike pair distributions (Supplementary Vid. 1). A novel algorithm extracted isolated clusters from single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the dataset \cite{reyes2014laminar}. It was also shown, however, that first spike pairs within these isolated clusters were positively correlated above chance-levels. These correlated clusters represent reliable regions of response distribution space that allow analysis of short- and long-timescale spike timing noise correlations. A large proportion of these noise correlations would have been missed without clustering.

Interestingly, many of the correlated clusters had non-unitary (different from $45^{\circ}$) correlation angles (including the green cluster in Fig. X).
Importantly, many clusters remained correlated with non-unitary gradient after controlling for longer-timescale processes (including neural adaptation). Such non-unitary gradient correlations are indicative of a short-timescale fluctuating state varying the spike times of two neurons at a different rate, such that the relative difference between the spike times varies with respect to this state.
This cannot explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle).

This novel finding is referred to as `adaptation-independent clustered time-warp spike time noise correlations'. Here, time-warp refers to the finding that pairwise relative spike time differences vary with respect to a state.
% Particularly, within cluster spike times vary at different rates w.r.t. an underlying state, such that relative differences vary with res
% by some short-timescale fluctuating continuous cortical state. 
The total number of low-dimensional states underlying the correlative structure of different pairs across the population, however, must be confirmed. 
% This phenomenon is referred to as `adaptation-independent clustered time-warped spike-timing noise correlations'. 
The findings differ from previous `time-warp' studies, which aim to align spike trains over trials and do not simultaneously consider potential encoding capacity, clustering, adaptation and different warping functions for different neurons \cite{williams2020discovering}.

This characterisation is useful to the fixed-state decoder interpretation. More interestingly, these noise correlations open the possibility of state-dependent spike-time coding. Noise correlations were found for pairs of neurons separated over multiple barrels and cortical layers, giving evidence of a spatially distanced short-timescale fluctuating state, which could also cause downstream neurons to respond in a state-dependent manner. Analysis estimated state-dependent relative first spike difference response distributions for each adaptation-controlled correlated cluster, giving a prediction of the capacity of state-dependent encoding (based on distribution size), which surpasses previously proposed coding forms.

% Whilst stimulus-dependency of noise correlations is not investigated in this study, estimates of potential encoding capacity are made by analysis of state-dependent response distribution size.

@@

% Using a large dataset of spike sorted recordings in anesthetized rat barrel cortex, the paper aims to address the following questions:

@@

Analysing the joint responses of two or more units to repeated presentations of a single stimulus offers the opportunity to discern noise correlations caused by adaptative or non-adaptative processes.
A two-stage algorithm was developed to quantify the occurrence and angle of correlated clusters observed in first spike response distributions for pairs of units (Fig. X; top left of Supplementary Vid. 1). Stage 1 of the algorithm extracted well isolated clusters defined by the $4\sigma$ boundary of flat 2D Gaussian ellipses (see Methods \ref{methods:flat_clusters}; e.g. flat solid green ellipse in Fig. 3a).
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0}
The algorithm was verified by confirming that the algorithm extracted correlated clusters at chance levels for multiple control datasets (see Methods \ref{methods:correlation_surrogate_control_datasets}; Supplementary Fig. 3). The correlated clusters can be analysed to test whether the correlations are explained by adaptative or non-adaptative processes.

@@

\subsection*{Physiological Relevance}
% It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. 
Testing the state-dependent decoding performance of the clusters is left for future studies, as it would require explicit knowledge of the underlying shared state or an estimate from the spike times of other neurons (which is difficult for the current extracted cluster counts per stimulus in single experimental sessions).
Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by low trial overlap between clusters. 
Neuronal subthreshold membrane potentials would be a candidate for such state, as they are correlated between neurons across areas \cite{poulet2008internal} and affect action potential initiation \cite{petersen2003interaction}. The LFP may offer a useful proxy for the subthreshold membrane potential.

This paper provides a useful step towards investigating this and shows that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.

@@
